Inference in Hidden Markov Models with Explicit State Duration Distributions

### 2011-12-20 tex/abstract.tex
my final (I hope) edits
Frank's last (hopefully) changes

Explicit-state-duration hidden Markov models (EDHMM) are HMMs that have latent states consisting of both discrete state-indicator and discrete state-duration random variables.  In contrast to the implicit geometric state duration distribution possessed by the standard HMM, EDHMMs allow the direct parameterisation and estimation of per-state duration distributions. As most duration distributions have support on all positive integers, truncation or other approximations are usually required to perform EDHMM inference.  In this letter we borrow from the inference techniques developed for unbounded state-cardinality (nonparametric) variants of the HMM and use them to develop a tuning-parameter free inference approach for EDHMMs.  We illustrate the performance gains of this approach using synthetic data.

>>>>

Explicit-state-duration hidden Markov models (EDHMM) are HMMs that have latent states consisting of both discrete state-indicator and discrete state-duration random variables.  In contrast to the implicit geometric state duration distribution possessed by the standard HMM, EDHMMs allow the direct parameterisation and estimation of per-state duration distributions. As most duration distributions are defined over the positive integers, truncation or other approximations are usually required to perform EDHMM inference.  In this letter we borrow from the inference techniques developed for unbounded state-cardinality (nonparametric) variants of the HMM and use them to develop a tuning-parameter free, black-box inference procedure for EDHMMs.  %We illustrate the performance gains of this approach using synthetic data.

### 2011-12-20 tex/introduction.tex
Frank's last (hopefully) changes

The explicit state duration HMM (EDHMM) \cite{Rabiner89}, a member of the Hidden Semi Markov Models \cite{Yu10}, was developed to 
allow explicit parameterization and direct inference of state duration 
distributions.   EDHMM estimation and inference can be performed using the 
forward-backward algorithm; specifically if the sequence is short or a tight 
``allowable'' duration interval for each state is hard-coded a priori 
\cite{Yu2006}.   If the sequence is short then forward-backward can be run 
on a state representation that allows for all possible durations up to the 
observed sequence length.  If the sequence is long then forward-backward 
only remains computationally tractable if only transitions between durations 
that lie within pre-specified allowable intervals are considered.   If the 
true state durations lie outside those intervals then the resulting model 
estimates will be incorrect: the learned duration distributions will reflect 
only what is allowed given the pre-specified duration intervals. % (which may poorly describe the true duration distributions).

>>>>

The EDHMM  (or, equivalently, the hidden semi-Markov model \cite{Yu10}) was developed to 
allow explicit parameterization and direct inference of state duration 
distributions.   EDHMM estimation and inference can be performed using the 
forward-backward algorithm; though only if the sequence is short or a tight 
``allowable'' duration interval for each state is hard-coded a priori 
\cite{Yu2006}.   If the sequence is short then forward-backward can be run 
on a state representation that allows for all possible durations up to the 
observed sequence length.  If the sequence is long then forward-backward 
only remains computationally tractable if only transitions between durations 
that lie within pre-specified allowable intervals are considered.   If the 
true state durations lie outside those intervals then the resulting model 
estimates will be incorrect: the learned duration distributions can only reflect 
 what is allowed given the pre-specified duration intervals. % (which may poorly describe the true duration distributions).

### 2011-12-20 tex/results.tex
Frank's last (hopefully) changes

\caption{Mean number of transitions considered per time point by the beam sampler for 1000 post-burn-in iterations on data from Figure \ref{fig:experiment1_results}. Consider this in comparison to the $(KT)^2 = O(10^6)$ transitions that would need to be considered by standard forward backward without truncation, the only surely safe, truncation-free alternative.}
\label{fig:allowed}
\end{figure}

>>>>

\caption{Mean number of transitions considered per time point by the beam sampler for 1000 post-burn-in sweeps on data from Figure \ref{fig:experiment1_results}. Consider this in comparison to the $(KT)^2 = O(10^6)$ per time point transitions that would need to be considered by standard forward backward without truncation, a surely-safe, truncation-free, but computationally impractical alternative.}
\label{fig:allowed}
\end{figure}

### 2011-12-20 tex/sampling.tex
Frank's last (hopefully) changes

Our aim is to estimate the conditional posterior distribution of the latent states ($\mathcal{X}$ and $\mathcal{D}$) and parameters ($\theta, \lambda$ and $A$) given observations $\mathcal{Y}$ by samples drawn via Markov chain Monte Carlo. Sampling $\theta$ and $A$ given $\mathcal{X}$ proceeds per usual textbook approaches \cite{Bishop06}.  Sampling $\lambda$ given $\mathcal{D}$ is straightforward in most situations.  Gibbs sampling  $\mathcal{X}$ is possible but, for reasons similar to those in the case of graphical models with chain dependency structure, one would not expect Gibbs sampling to work well for this model.  The main contribution of this paper is to show how to generate posterior samples of  $\mathcal{X}$ and $\mathcal{D}$ efficiently given all other random variables.

@@

Unfortunately the sum in \eqref{eqn:forward recursion} has an infinite number of terms in the usual case of  duration distributions with countably infinite support. The standard approach to EDHMM inference involves truncating these sums in order to make them tractable. 
%If all durations are truncated at a maximum $d_\mathrm{max}$, yet many states are of duration far less than $d_{\mathrm{max}}$, this necessarily leads to much unnecessary computation
% if not all states have durations near $d_\mathrm{max}$. 
A reasonable computational artifice is to truncate all durations at some maximum $d_\mathrm{max}$, or to consider all possible durations up to the observed length of the sequence. This leads to per sample, forward-backward computational complexity of $O(T(Kd_\mathrm{max})^2)$  and  $O(T(KT)^2)$ respectively.  Alternatively one could risk inference that will simply fail if an actual duration lies outside hard-coded allowable duration intervals $d_\mathrm{min}<d<d_\mathrm{max}$.  This yields $O(T(K(d_\mathrm{max}-d_\mathrm{min}))^2)$ complexity.
The beam-sampler we propose behaves like a dynamic version of the latter, automatically scaling the allowable per-state duration intervals.   Its use of slice-sampling auxiliary variables results in an asymptotically exact sample with no risk of incorrect inference resulting from incorrectly pre-specified duration bounds.   We do not characterize the computational complexity of the proposed beam sampler in this work but note that it is upper bounded by $O(T(KT)^2)$ (i.e., the beam sampler admits durations of length equal to the entire sequence) but in practice is found to be as  or more efficient than the doubly hard-coded, risky approach.

>>>>

Our aim is to estimate the conditional posterior distribution of the latent states ($\mathcal{X}$ and $\mathcal{D}$) and parameters ($\theta, \lambda$ and $A$) given observations $\mathcal{Y}$ by samples drawn via Markov chain Monte Carlo. Sampling $\theta$ and $A$ given $\mathcal{X}$ proceeds per usual textbook approaches \cite{Bishop06}.  Sampling $\lambda$ given $\mathcal{D}$ is straightforward in most situations.  Indirect Gibbs sampling  of $\mathcal{X}$ is possible using auxiliary state-change indicator variables, but for reasons similar to those in \cite{Goldwater2009}, such a sampler will not mix well.  The main contribution of this paper is to show how to generate posterior samples of  $\mathcal{X}$ and $\mathcal{D}$.

@@

Unfortunately the sum in \eqref{eqn:forward recursion} has at worst an infinite number of terms in the case of  duration distributions with countably infinite support and at best a very large number of terms in the case of long sequences. The standard approach to EDHMM inference involves truncating considered durations to only those that lie between $d_\mathrm{min}$ and $d_\mathrm{max}$ or computation involving all possible durations up to the observed length of the sequence ($d_\mathrm{min}=0, d_\mathrm{max}=T$). This leads to per-sample, forward-backward computational complexity of $O(T(K(d_\mathrm{max}-d_\mathrm{min}))^2)$.
Truncation yields inference that will simply fail if an actual duration lies outside hard-coded allowable duration intervals.   Considering all possible durations up to length $T$ is often computationally impossible.
%This yields $O(T(K(d_\mathrm{max}-d_\mathrm{min}))^2)$ complexity.
The beam-sampler we propose behaves like a dynamic version of the truncation approach, automatically defining and scaling per-state duration truncation intervals.   Better though, the way it does this results in an asymptotically exact sample with no risk of incorrect inference resulting from incorrectly pre-specified duration truncations.   We do not characterize the computational complexity of the proposed beam sampler in this work but note that it is upper bounded by $O(T(KT)^2)$ (i.e., the beam sampler admits durations of length equal to the entire sequence) but in practice is found to be as  or more efficient than the risky hard-truncation approach.
